---
title: "class18"
author: "Siddharth (A18594442)"
format: pdf
---

# 1. Investigating pertussis cases by year

## Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
# had trouble with the datapasta package so went without it. 
cdc <- read.csv("U.S. Reported Pertussis Cases_ 1922 - 2025.csv")
#issue with (,) commasin data.  made collumn a character instead of an int causing it to not be recognized and breaking the graph. deleted commas in that collumn with gsub.

cdc$Number.of.Reported.Pertussis.Cases <- as.numeric(
  gsub(",", "", cdc$Number.of.Reported.Pertussis.Cases)
)
library(ggplot2)
ggplot(cdc) +
  aes(x = Year, y = Number.of.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(
    title = "Reported Pertussis Cases in the United States",
    x = "Year",
    y = "Number of Cases"
  )
```

# 2. A tale of two vaccines (wP & aP)

## Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

Between the wP vaccine and the aP vaccine there has been a plateau of cases, wtih new vases re-emerging since the introduction of the aP vaccine.


```{r}
ggplot(cdc) +
  aes(x = Year,
      y = Number.of.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red") +
  labs(
    title = "Reported Pertussis Cases in the US",
    x = "Year",
    y = "Number of Cases"
  )

```


## Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

The introduction of the aP vaccine has led to a ressurgence of casses of purtussis. This could be due to vaccine hesitancy, or resistance to the vaccine caused by evolution of the pathogen, and therefore lower protection afforded by the vaccine. 


#3. Exploring CMI-PB data


```{r}
# Allows us to read, write and process JSON data
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```


##Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

There are 87 acellular pertussis and 85 whole-cell pertussis vaccines in the dataset. 

```{r}
table(subject$infancy_vac)
```


##Q5. How many Male and Female subjects/patients are in the dataset?

There are 112 Female and 60 Male subjects/patients in the dataset. 

```{r}
table(subject$biological_sex)
```


##Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

##Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

(i)  the avg age of wP idividuals is 35 years old
(ii) the avg age of aP individuals is 28 years old 
(iii) the avg age of wP individuals is significantly different from the avg age of aP individuals
```{r}
library(lubridate)
subject$age <- today() - ymd(subject$year_of_birth)
library(dplyr)
#calculating the age of wp individuals
wp <- subject %>%
  filter(infancy_vac == "wP")
print("wp")
round(summary(time_length(wp$age, "years")))
# calculate the avg age of ap individuals 
ap <- subject %>%
  filter(infancy_vac == "aP")
print("ap")
round(summary(time_length(ap$age, "years")))
#using a t test to see if the ages are stat. sig. different
t.test(
  time_length(ap$age, "years"),
  time_length(wp$age, "years")
)
```


##Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
#contains the age at boost for all individuals
head(age_at_boost)
```

##Q9. With the help of a faceted boxplot or histogram, do you think these two groups are significantly different?

Yes, the two groups apperar to be significantly different. There is a grouping of much younger people for aP and a much more spread out but generally above 30 year old group of people for wP.


```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```


#Joining multiple tables
```{r}
library(jsonlite)

specimen <- read_json(
  "https://www.cmi-pb.org/api/specimen",
  simplifyVector = TRUE
)

# the only availible api from cmi-pb was plasma_ab_titer. the ab_titer api threw a 404 not found error.
titer <- read_json(
  "https://www.cmi-pb.org/api/plasma_ab_titer",
  simplifyVector = TRUE
)
# mapped data to objects
# use leftjoin to join 
meta <- left_join( specimen, subject)
dim(meta)
head(meta)

```


##Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.
```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

##Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

##Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

The most recent dataset, 2023, has 5,670 rows, it is the smallest. 

```{r}
table(abdata$dataset)
```

#4. Examine IgG Ab titer levels


```{r}
# taking imunoglobulin g only: 
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```


##Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0, 75) +
  facet_wrap(vars(visit), nrow = 2)
```

##Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PRN (Pertactin), FIM 2/3 (Fimbriae), FHA (filamentous hemagglutinin) all are antigens that are from the pertusis bacterium. As opposed to an antigen like OVA (ovalbumin) which comes from a chicken egg white protein, PRN,FIM 2/3, and FHA all have increasing antibody titers as time goes on indicating that the vaccine makes antibodies specific to pertussis. 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```


```{r}
# same plot with infancy vacc
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

##Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

I chose TT which I assume is Tuberculosis Toxin and compared it to FHA which is filamentous hemagglutinin from the bacterium B. pertussis


```{r}
filter(igg, antigen=="TT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
```{r}
filter(igg, antigen=="FHA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()

```

##Q16. What do you notice about these two antigens time courses and the PT data in particular?

I notice that the FHA data shows a gradual increase with distinct peak around visit 7 and then a decline, while the TT antigen ag titer is more variable and looks like noise with no real increase or decrease as time goes on. 



##Q17. Do you see any clear difference in aP vs. wP responses?

I do not see a clear difference between aP and wP responses, it seams as though wP might have a more gradual ab titer increase betweens week 1 and 4 but afterwards levels seem to be the same between aP and wP. 


```{r}
# this is for 2021 data

abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")

```

##Q18. Does this trend look similar for the 2020 dataset?

The overall trend looks to be the same but there is clearly more data for the 2020 dataset extending to 400 days for the wP vaccine. 

```{r}
# this is for 2022 data
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG", antigen == "PT") %>%
  ggplot() +
    aes(x = planned_day_relative_to_boost,
        y = MFI_normalised,
        col = infancy_vac,
        group = subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 14, linetype = "dashed") +
    labs(title = "2020 dataset IgG PT",
         subtitle = "Dashed lines indicate day 0 and day 14")
```


#5. Obtaining CMI-PB RNASeq data

```{r}

url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"
rna <- read_json(url, simplifyVector = TRUE) 
# previously meta <- inner_join(specimen, subject)
#join the rna expression data with our metadata meta
ssrna <- inner_join(rna, meta)
```

##Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}

```
















