---
title: "class11"
author: "Siddharth (A18594442)"
format: pdf
---
#set pathing and read in a pdb file
```{r}
library(bio3d)
pdb <- read.pdb("FindageneprojectCMSS1homolog_23c00/FindageneprojectCMSS1homolog_23c00_unrelaxed_rank_001_alphafold2_ptm_model_1_seed_000.pdb")
pdb
```

#Make a vector of input PDB file names that we can read into R 
```{r}
# Change this for YOUR results dir name
results_dir <- "hivprdimer_23119/" 
pdb_files <- list.files("FindageneprojectCMSS1homolog_23c00/", pattern= ".pdb", full.names = TRUE)
pdb_files
```

```{r}

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")

```

```{r}
# view sequences
pdbs

```

```{r}
#rmsd gets the structural distance between the sets of coordinates 
rd <- rmsd(pdbs, fit=T)
range(rd)
```

```{r}
# drawing a heatmap of these RMSD matrix values

library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```
```{r}
# Read a reference PDB structure
pdb <- read.pdb("1hsg")

```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```
```{r}
core <- core.find(pdbs)
```

```{r}

core.inds <- print(core, vol=0.5)

```


```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")

```


```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```
```{r}
# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```


```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```















#Citations
Baek, Minkyung, Frank DiMaio, Ivan Anishchenko, Justas Dauparas, Sergey Ovchinnikov, Gyu Rie Lee, Jue Wang, et al. 2021. “Accurate Prediction of Protein Structures and Interactions Using a Three-Track Neural Network.” Science 373 (6557): 871–76. https://doi.org/10.1126/science.abj8754.

Jumper, John, Richard Evans, Alexander Pritzel, Tim Green, Michael Figurnov, Olaf Ronneberger, Kathryn Tunyasuvunakool, et al. 2021. “Highly Accurate Protein Structure Prediction with AlphaFold.” Nature 596 (7873): 583–89. https://doi.org/10.1038/s41586-021-03819-2.

Mirdita, Milot, Konstantin Schütze, Yoshitaka Moriwaki, Lim Heo, Sergey Ovchinnikov, and Martin Steinegger. 2022. “ColabFold: Making Protein Folding Accessible to All.” Nature Methods 19 (6): 679–82. https://doi.org/10.1038/s41592-022-01488-1.

