---
title: "class10"
author: "Siddharth (A18594442)"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

###1 read.pdb() ###2 NMA() ###3 view.nma() ; mktrj() ; plot()

#Q1: What percentage of structures in the PDB are solved by X-Ray and
Electron Microscopy.

XRay- 80.95% EM- 12.84%

```{r}
pdb <- read.csv("pdb_stats.csv")
head(pdb)
```

```{r}
# total number of structures
total_structures <- sum(pdb$Total)

# totals by method
xray_total <- sum(pdb$X.ray)
em_total <- sum(pdb$EM)

# percentages
xray_percent <- (xray_total / total_structures) * 100
em_percent <- (em_total / total_structures) * 100

xray_percent
em_percent
```

#Q2: What proportion of structures in the PDB are protein?

97.92 percent of structures in the PDB are protein

```{r}
# grab the total num of protein like structures
protein_total <- sum(
  pdb$Total[pdb$Molecular.Type %in% c(
    "Protein (only)",
    "Protein/Oligosaccharide",
    "Protein/NA"
  )]
)
# calculate protien percentage 
protein_percent <- (protein_total / total_structures) * 100

protein_percent
```

#Q3: Type HIV in the PDB website search box on the home page and
determine how many HIV-1 protease structures are in the current PDB?

There are 1,227 HIV-1 protease structures in the current PDB

#Q4: Water molecules normally have 3 atoms. Why do we see just one atom
per water molecule in this structure?

PDB structures can only resolve the oxygen atom in H2O and so therefore
only see one atom in the molecule. Hydrogen atoms scatter Xrays weakly
for Xray crystalography techniques.

# Q5: There is a critical “conserved” water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have

The water molecule in the binding site is HOH 332, it looks like it
participates in hydrogen bonding with the active site.

#Q6: Generate and save a figure clearly showing the two distinct chains
of HIV-protease along with the ligand. You might also consider showing
the catalytic residues ASP 25 in each chain and the critical water (we
recommend “Ball & Stick” for these side-chains). Add this figure to your
Quarto document.

1\.

![Critical water molecule HOH 332 in the active site of HIV-1
protease](images/1HSG.png)

2\.

![Highlighted in orange, ASP 25 (D25) aspartic acid residue of the HIV-1
protease B chain](images/1HSG%20(1).png)

3.![Highlighted in green, ASP25 (D25) aspartic acid residue of the HIV-1
protease A chain](images/1HSG%20(2).png)

Highlighted in green, ASP25 (D25) aspartic acid residue of the HIV-1
protease A chain

Steps for 4. 1 read.pdb() 2 NMA() 3 view.nma() ; mktrj() ; plot()

```{r}
library(bio3d)
library(NGLVieweR)
library(msa)
```

```{r}
pdb <- read.pdb("1hsg")
```

#Q7: How many amino acid residues are there in this pdb object?

There are 198 amino acid residues in the pdb object.

```{r}
pdb

```

#Q8: Name one of the two non-protein residues?

MK1 is one of the two non protien residues.

#Q9: How many protein chains are in this structure?

There are 2 protien chains in this structure with "values" A and B.

## skipped the in between parts due to incompatibility with rendering to PDF.

#5. Comparitive structural analysis of ADK family 

1. get.seq() - input :database identifier, output: protien sequence 
2. blast.pdb()-\>get.pdb() input: fasta output: finds and downloads all related structure FASTAs 
3. pbd.align()

```{r}
# this is just to give context to the pakages Q10  is talking about:

#install.packages("bio3d")
#install.packages("NGLVieweR")

#install.packages("remotes")
#remotes::install_github("bioboot/bio3dview")

#install.packages("BiocManager")
#BiocManager::install("msa")

```

#Q10. Which of the packages above is found only on BioConductor and not CRAN?

the msa pakage is only found in BioConductor and not CRAN.

#Q11. Which of the above packages is not found on BioConductor or CRAN?:

bio3dview is the pakage that is not found on BioConductor or CRAN.

#Q12. True or False? Functions from the pak package can be used to install packages from GitHub and BitBucket?

TRUE, functions from the pak pakage can be used to install pakages from GitHub and BitBucket. 

##Search and retrieve ADK structures

```{r}
aa <- get.seq("1ake_A")
```

```{r}
aa
```

##Q13. How many amino acids are in this sequence, i.e. how long is this
sequence?

There are 214 amino acids in this sequence.

```{r}
# blast the aa sequence to the ncbi database 
b <- blast.pdb(aa)
```

```{r}
hits <- plot(b)
```

```{r}
head(hits$pdb.id)
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")

```

Quick interactive structural view

```{r}

#library(bio3dview)

#view.pdbs(pdbs)
```

PCA of all this structural data (x, y, and z atom coordinates)

```{r}
# pass this to pca 

pc.xray <- pca(pdbs)
plot(pc.xray)

```

```{r}
#conformer plot of the conformational variability within all the related PDB structures. 
plot(pc.xray, 1:2)
```

```{r}
#view.pca(pc.xray)
# gives a superimposed image of all BLAST hit structures
```

Interactive view of the pc1 captured structural differences

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")

```

```{r}
# gives the x y and z coordinates for wach atom in the structure
pdbs$xyz
```
