---
title: "class06"
author: "Siddharth Ranade"
format: pdf
---

##Q1. Write your first R function
```{r}
add <- function(x,y)
{
  result <- x+y
  return(result)
}
add(4,7)
```

add function V2.

```{r}
# input is a
addV2 <- function(x, y = NULL) {
  if (is.null(y)) {
    # If only one input is provided, sum the elements of that input
    return(sum(x))
  } else {
    # If two inputs are provided, sum them together
    return(x + y)
  }
}
addV2(c(4,7,10))
addV2(4,7)
```

add function V3.

```{r}
addV3 <- function(...)
{
  return(sum(...))
}
addV3(1, 2, 3, -4)
```

##Q2 Write a `generate_dna()` function

#A. V1 `?sample` shows me that `sample()` takes a sample of the specified size from the elements of x using either with or without replacement.

```{r}
nucleotides <- c("A","T","G","C")
generate_dnaV1 <- function(x)
{
  return(sample(nucleotides,x, replace=TRUE))
}
generate_dnaV1(6)
```

#B.V2 ?paste and ?paste0 tells me that I can concatenate vectors after converting to character. Concatenation happens in two basically different ways, determined by collapse being a string or not.

```{r}
nucleotides <- c("A","T","G","C")
#takes x=length and y= boolean for true false if you want a single element, if they pass TRUE we return a single element 
nucleotides <- c("A", "T", "G", "C")

generate_dnaV2 <- function(x, y = FALSE) {
  dna <- sample(nucleotides, x, replace = TRUE)
  #collapse and paste0 concatenates the individual strings into one long string 
  if (y) {
    return(paste0(dna, collapse = ""))
  } else {
    return(dna)
  }
}

generate_dnaV2(6,TRUE)
generate_dnaV2(6,FALSE)
```

#C. will generate a FASTA format sequence with id line ?cat

```{r}
nucleotides <- c("A","T","G","C")

generate_dna_fasta <- function(x){

  dna <- paste0(
    sample(nucleotides, x, replace = TRUE),
    collapse = ""
  )
# \n is a new line escape,
  cat(">len", x, "\n", dna, "\n", sep="")
}

generate_dna_fasta(6)
```

##Q3. Write a generate_protein() function

```{r}
protiens <- c("A", "R", "N", "D", "C", "E", "Q", "G", "H", "I", "L", "K", "M", "F", "P", "S",
"T", "W", "Y", "V")
generate_protien <- function(x){
  return(sample(protiens, x ,replace = TRUE))
}
generate_protien(6)
```

##Q4. Generate random protein sequences of length 6 to 13

```{r}
proteins <- c("A","R","N","D","C","E","Q","G","H","I",
              "L","K","M","F","P","S","T","W","Y","V")

# Function: generate only one protein sequence of length x where x is the 
generate_protein <- function(x){
  paste0(
    sample(proteins, x, replace = TRUE),
    collapse = ""
  )
}

# Loop from 6 to 13 and print FASTA
for(i in 6:13){
  cat(
    paste0(
      ">id.", i, "\n",
      generate_protein(i), "\n"
    )
  )
}

```

##Q5. BLASTp search against nr — are your peptides “unique in nature”?

| Length (aa) | Best hit % identity | Best hit % coverage | Unique? (Y/N) |
|-------------|---------------------|---------------------|---------------|
| 6           | 100%                | 100%                | N             |
| 7           | 100%                | 100%                | N             |
| 8           | 100%                | 100%                | N             |
| 9           | 88.89%              | 100%                | Y             |
| 10          | 100%                | 90%                 | Y             |
| 11          | 90%                 | 91%                 | Y             |
| 12          | 81.82%              | 92%                 | Y             |
| 13          | 83.33%              | 92%                 | Y             |

>a. At which sequence length do your randomly generated peptides start to look “unique in nature”
>(i.e. no 100% coverage + 100% identity hit)? \[1 pt\]

These peptides start to look "unique in nature" once we hit around 9 peptides in length, at that point it seems that the length of the chain is long enough to distinguish itself as "unique".



>b. Speculate why very short random peptides are almost always found in nr, while longer ones
typically are not. Your answer should refer both to the size of the sequence space (20𝐿 for a
peptide of length 𝐿) and to the size of the known protein universe. \[1 pt\]

Short peptides are more likely to be a match because of the fact that there are only 20 amino acids, when given a short sequence, the probability that the sequence is included in another "natural" one is quite high with that probability dropping off as you increase the length of that string of amino acids. 


## Q6.Connecting your findings to immunology (MHC class II and T-cell activation)

>In 3–6 sentences total and using your Q5 data and the reasoning from Q5b, what do you think
this minimum length is and why might it be a bad design choice for the immune system to present
very short peptides? [3 pt]

The minimum length of peptide that can be loaded onto class II MHC is likely to be around 11 peptides as the data in Q5 shows that past 10 peptides we start to lose perfect coverage and identity. More than likely the true length is longer than 11 peptides so that the immune system does not activate against self peptides mistakenly. It would be a bad design choice for the immune system to present very short peptides because they would be too nonspecific as shown in the data from Q5, peptides shorter than 9 amino acids were all "non unique", and could potentially come from self antigen causing autoreactivity. 



