---
title: "Class 5: Data viz with ggplot2"
author: "Siddharth Ranade (PID: A18594442)"
format: pdf
toc: true
---

## Background
There are many graphics systems for making plots and igres. These inclide so-called *"base R"graphics* like the `plot()` function and add on packages like ggplot2

We can use. the in-built `cars` dataset

```{r}
plot(cars)
```
The main fnction in the **ggplot2** pakage is called ggplot2 
Before I can use ggplot2 I need to install it in my computer. To do this we can use the unction `install.packages("ggplot2")`
> **N.B.** Never use this in a code chunk because it will needelessly re install each time the code chunk is ran. 

```{r}
library(ggplot2)
```
```{r}
ggplot(cars)
```
Every ggplot has at least 3 layers:
- **data** (data frame of stuff we want to put in the plot)
- **aes** aesthetics (how the data maps to the plot)
- **geom** geometry layer (how you want the plot to be drawn, points lines...)

ex. ggplot()+ aes()+geom_"function"()

```{r}
ggplot(cars)+ aes(x=speed, y=dist)+geom_point()
```
## Adding some custom features 
Lets add a trend line that shows the relationship between speed and distance
```{r}
ggplot(cars)+
aes(x=speed, y=dist)+
geom_point()+
geom_smooth(method = "lm",se=FALSE)+
theme_bw()+
labs(title = "Stopping Distance Of Old Cars", x="Speed (MPH)", y="Distance (Ft)")
                                                                          

```
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
sum(genes$State=="up")

```
A useful new function in this context is the `table function`:
```{r}
table(genes$State)
```
## Creating Scatter plots
```{r}
  ggplot(genes)+ 
  aes(Condition1,Condition2, col=State)+
  geom_point()+
  scale_colour_manual(values=c("blue","gray","red") )+
  labs(title="Gene Expression Change With Drug Treatment", x= "No drug", y="drug")

```
## Going Further
Here we read in the famous gapminder dataset
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```
>Q how many total entries (ie. rows) are there in this dataset?

```{r}
nrow(gapminder)
```

>Q how many countries are in this dataset?

```{r}
length(unique(gapminder$country))
```
Lets make a plot of the entire dataset: 
Plot of "gdpPercap" vs"lifeExp" colored by "continent"
```{r}
p <- ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col= continent)+
  geom_point(alpha=0.3)
```
I can add more layers to `plot`
```{r}
p + facet_wrap(~continent)
```
Make a plot for 1997 and 2007 
```{r, message=FALSE}
library(dplyr)
gap2007 <- filter(gapminder, year==2007)
gap1997 <- filter(gapminder,year==1997)
gap2007or1997 <- filter(gapminder, year==2007|1997)
p2007 <- ggplot(gap2007)+
  aes(gdpPercap, lifeExp, col= continent, size= pop)+
  geom_point()+ labs(title= "GDP vs Life Expectancy in 2007")
p2007
p1997 <- ggplot(gap1997)+
  aes(gdpPercap, lifeExp, col= continent,size= pop)+
  geom_point()+ labs(title= "GDP vs Life Expectancy in 1997")
p1997
p2007or1997 <-  ggplot(gap2007or1997)+
  aes(gdpPercap, lifeExp, col= continent,size= pop)+
  geom_point(alpha=0.3)+ labs(title= "GDP vs Life Expectancy in 1997 and 2007")
p2007or1997
```
Histogram of Life Expectancy
>Colored by continent
>Faceted by continent

```{r}
ggplot(gapminder)+
  aes(lifeExp, col=continent, fill=continent)+
  geom_histogram()

ggplot(gapminder)+
  aes(lifeExp,col=continent, fill=continent)+
  geom_histogram()+
  facet_wrap(~continent)
  
```

