---
title: "class07"
author: "Siddharth Ranade (PID: A18594442)"
format: pdf
---

# PCA of UK food data

##importing data
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```
##Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

I can use `ncol()` to get the number of columns in the data frome. There are 5 columns in the data frame. I can also use the `dim()` function

```{r}
ncol(x)
dim(x)
```
```{r}
#View(x)
```
removing the 5th collumn which we dont want, we just want it as the header for the data
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```
```{r}
dim(x)
```
##Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the method of using `x <- read.csv(url, row.names=1)` as it is much more clean and efficient, as well as less confusing and prone to error while looking at the header.

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))

```

#Q3: Changing what optional argument in the above barplot() function results in the following plot?

Making the beside argument false changes the barplot to be stacked. 

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))

```
#Making data format long 
```{r}
library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```


```{r}
library(ggplot2)
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()

```

##Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

Changing the `geom_col()` argument of position to "stack" results in a stacked bar plot. 

```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

##Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

For a given comparison in the pairwise plot, if a point lies on a perfect diagonal line where x=y, then the values for both the comparison variables are the same. This is to say that for a comparison between two countries if a given food item lies on the x y axis then the consumption of that food is the same in both countries. 

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )

```

##Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on the clustering on the heat map, foods seem to be paired together based on their consumption level. For example soft drinks ans cereals cluster together based on their relatively similar consumption level. Some differences I see are that Northern Ireland consumes more potatoes than other UK countries. We also see that England, Wales and Scotland cluster closer together when it comes to consumption habits. 

#PCA to the rescue: 
```{r}
pca <- prcomp( t(x) )
summary(pca)

```
##Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()


```

##Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.


```{r}
p <- ggplot(df) +
  aes(x = PC1, y = PC2, label = Country, color = Country) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()

# add colors
p + scale_color_manual(values = c(
  "England" = "yellow",
  "Scotland" = "blue",
  "Wales" = "red",
  "N.Ireland" = "darkgreen"
))
```

```{r}
z <- summary(pca)
z$importance

```
#SCREE plot
```{r}
# taking the square of the standard deviation gives us the variance; from the example code
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```
#Loading plot for PC1

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

```
## Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

The two food groups that impact PC2 the most are soft drinks and alcoholic drinks. PC2 mainly tells us about the drinks that are consumed by a country. A point's position representing a country on the PC2 axis is most influenced by soft drink and alcoholic drink consumption. 

```{r}
#replaced PC1 var with PC2
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))

```


#Explaining the differences between: PC Score Plot, Loading Plot, and a SCREE Plot

The difference between a PC score plot, a loading plot and a SCREE plot are the ways in which they examine PCA. A loading plot looks at the way the "multiple dimensions", which have been condensed to one axis/vector, contribute to the value of a variable along the PCA's axis. In doing so the loading plot reveals the way in which certain factors influence the PCA. A PC Score plot looks at principal components against each other; reducing multidimensional data into 2d. Points along a PC score plot are related by their distance from one another, the closer they are the more related they are. A SCREE plot is a plot that looks at how much of the variation in the data set is explained by a given PCA. The SCREE plot is a measure of the PCA and its ability to capture or explain the difference between variables in a dataset. 

